---
title: "HumanTranscriptomeCompendium Illustration: Hepatocellular Carcinoma Tumor/Normal/Thrombosis"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Working with htxapp: Tumor/normal/portal vein tumor thrombosis transcriptomes in Hepatocellular carcinoma}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
bibliography: htxbib.bib
---

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
library(BiocStyle)
library(htxapp)
library(restfulSE)
library(HumanTranscriptomeCompendium)
library(ggplot2)
library(table1)
})
```

# Introduction

We used the app at https://vjcitn.shinyapps.io/cancer9k 
to extract SRA study SRP069212.  This dataset was
contributed to SRA in conjunction with
"Recurrently deregulated lncRNAs in hepatocellular
carcinoma", a Nature Communications paper 
(@Yang2017 PMID 28194035).
The data are saved as `pairedHCC.rds`.  Reads were requantified
using salmon (by Sean Davis of NCI) and moved into the
`/shared/bioconductor` folder of the HDF Scalable Data Service
at `http://hsdshdflab.hdfgroup.org`.

We saved the dataset using the cancer9k app download facility,
to a local file `pairedHCC.rds`.  We'll use this to understand
aspects of the design of the study, and to replicate aspects
of the published analysis.
```{r lkhcc}
library(restfulSE)
hccdat = readRDS("pairedHCC.rds")
hccdat
```

# Data exploration

The retrieved SummarizedExperiment instance has a `colData` with only
four columns.  The htxapp adds sample attribute information retrieved with SRAdbV2.
```{r lkmet}
dplyr::glimpse(metadata(hccdat)$sampleAttsSRP069212)
```

We add this table to the colData using the `bindSingleMetadata` function
of the `htxapp` package.
```{r add}
library(htxapp)
hccdat = bindSingleMetadata(hccdat)
```

## Sample characteristics

Now the `r CRANpkg("table1")` package can be applied to `colData`
to get an overview of sample characteristics.
Note that here the `n=` entries DO NOT refer to independent observations.

```{r lkt1,results="asis"}
table1(~source_name+hbv.infection+cirrhosis+age|gender,data=colData(hccdat))
```

We can see that this study involves up to three assays per participant
(tumor, adjacent normal, portal vein tumor thrombosis).  Fractions
of individuals with cirrhosis and HBV infection are also noteworthy.

## Assay metadata

We use the function `addRD` from HumanTranscriptomeCompendium to
add rowData derived from Gencode 27.
```{r addg27}
hccdat = addRD(hccdat)
dplyr::glimpse(as.data.frame(rowData(hccdat)))
```

## Assay

The assay quantifications are accessed via the `r Biocpkg("DelayedArray")`
protocol.

```{r lkass}
assay(hccdat)
```

For targeted inquiries, we do not need to 'download' the
complete expression data.  Here we will check on LINC01018.
```{r lkpl}
feature_trace("LINC01018", hccdat) + xlab(" ") + guides(colour="none")
```
This display uses quantifications distinct from those
used by the authors.  The picture is consistent with notion
that the distribution of abundance of LINC01018 differs
between samples from tumor and adjacent normal tissue.

## Further work

The full assay data can be retrieved and loaded into memory
using `m <- as.matrix(assay(hccdat)); assay(hccdat) = m`.  This
operation took a little more than a minute on a mediocre network.
Speed of retrieval will also depend on load at the HDF Scalable
Data Service.

# References
