---
title: "HumanTranscriptomeCompendium Illustration: Hepatocellular Carcinoma Tumor/Normal/Thrombosis"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Working with htxapp: Tumor/normal/thrombosis transcriptomes in Hepatocellular carcinoma}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
library(BiocStyle)
library(htxapp)
library(restfulSE)
library(HumanTranscriptomeCompendium)
library(ggplot2)
library(table1)
})
```

# Introduction

We used the app at https://vjcitn.shinyapps.io/cancer9k 
to extract SRA study SRP069212.  This dataset was
contributed to SRA in conjunction with
"Recurrently deregulated lncRNAs in hepatocellular
carcinoma", a 2017 Nature Communications paper by Yang
and colleagues from Tsinghua University (PMID 28194035).
The data are saved as `pairedHCC.rds`.  Reads were requantified
using salmon (by Sean Davis of NCI) and moved into the
`/shared/bioconductor` folder of the HDF Scalable Data Service
at `http://hsdshdflab.hdfgroup.org`.

We saved the dataset using the cancer9k app download facility,
to a local file `pairedHCC.rds`.  We'll use this to understand
aspects of the design of the study, and to replicate aspects
of the published analysis.
```{r lkhcc}
library(restfulSE)
hccdat = readRDS("pairedHCC.rds")
hccdat
```

# Data exploration

The retrieved SummarizedExperiment instance has a `colData` with only
four columns.  The htxapp adds sample attribute information retrieved with SRAdbV2.
```{r lkmet}
dplyr::glimpse(metadata(hccdat)$sampleAttsSRP069212)
```

We add this table to the colData using the `bindSingleMetadata` function
of the `htxapp` package.
```{r add}
library(htxapp)
hccdat = bindSingleMetadata(hccdat)
```

## Sample characteristics

Now the `r CRANpkg("table1")` package can be applied to `colData`
to get an overview of sample characteristics.
```{r lkt1,results="asis"}
table1(~source_name+hbv.infection+cirrhosis+age|gender,data=colData(hccdat))
```
We can see that this study involves up to three assays per participant
(tumor, adjacent normal, portal vein tumor thrombosis).  Fractions
of individuals with cirrhosis and HBV infection are also noteworthy.

## Assay metadata

We use the function `addRD` from HumanTranscriptomeCompendium to
add rowData derived from Gencode 27.
```{r addg27}
hccdat = addRD(hccdat)
dplyr::glimpse(as.data.frame(rowData(hccdat)))
```

## Assay

The assay quantifications are accessed via the `r Biocpkg("DelayedArray")`
protocol.

```{r lkass}
assay(hccdat)
```

For targeted inquiries, we do not need to 'download' the
complete expression data.  Here we will check on LINC01018.
```{r lkpl}
rownames(hccdat)[which(rowData(hccdat)$gene_name == "LINC01018")]
l1018 = as.numeric(assay(hccdat["ENSG00000250056.5",]))
dd = data.frame(l1018, colData(hccdat)[,c("patient.id", "source_name")])
library(ggplot2)
ggplot(dd, aes(y=log(l1018+1), x=source_name, group=patient.id)) + geom_point() + geom_path(aes(colour=patient.id))
```

